Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2022-11-07 nakayama/images/algebraic-structures.tex
Edits to scientific prose, comments added

\begin{tikzpicture}[->,>=stealth,shorten >=1pt,auto,node distance=1cm,semithick]
   \node[Snode] (Semigroup) {Semigroup};
   \node[Snode] (Monoid) [below = of Semigroup]{Monoid};
   \node[Snode] (Group) [below =of Monoid] {Group};
   \node[Snode] (AbelianGroup) [below =of Group] {Abelian Group};
   \node[Snode] (Subgroup) [left =1.3cm of AbelianGroup] {Subgroup};
   \node[Snode] (Ring) [below = of Subgroup] {Ring};
   \node[Snode] (CommutativeRing) [below =of Ring] {Commutative Ring};
   \node[Snode] (Ideal) [right =1.3 of CommutativeRing] {Ideal};
   \node[Snode] (Module) [right = 1.3cm of Ideal] {Module};
%   \node[node] (H1) [below = of G1] { };
%   \node[node] (H2) [below = of G2] { };
\path
(Semigroup) edge (Monoid)
(Monoid) edge (Group)
(Group) edge (AbelianGroup)
(Group) edge (Subgroup)
(AbelianGroup) edge (Ideal)
(AbelianGroup) edge (Ring)
(AbelianGroup) edge (Module)
(Ring) edge (CommutativeRing);
%(G1) edge (H1);
%(G2) edge (H2);
\end{tikzpicture}

>>>>

\begin{tikzpicture}[->,>=stealth,shorten >=1pt,auto,node distance=1cm,semithick]
   \node[Snode] (Semigroup) {Semigroup};
   \node[Snode] (Monoid) [below = of Semigroup]{Monoid};
   \node[Snode] (Group) [below =of Monoid] {Group};
   \node[Snode] (AbelianGroup) [below =of Group] {AbelianGroup};
   \node[Snode] (Subgroup) [left =1.3cm of AbelianGroup] {Subgroup};

   \node[Snode] (Ring) [below = of AbelianGroup] {Ring};
   \node[Snode] (CommutativeRing) [below =of Ring] {CommRing};
   \node[Snode] (Module) [right = 1.3cm of CommutativeRing] {Module};
   \node[Snode] (Submodule) [left = 1.0 of CommutativeRing] {Submodule};
   \node[Snode] (Ideal) [left =1.0 of Submodule] {Ideal};
   %   \node[node] (H1) [below = of G1] { };
%   \node[node] (H2) [below = of G2] { };
\path
(Semigroup) edge (Monoid)
(Monoid) edge (Group)
(Group) edge (AbelianGroup)
(Group) edge (Subgroup)
(Subgroup) edge (Ideal)
(AbelianGroup) edge (Ring)
(AbelianGroup) edge (Module)
(Ring) edge (CommutativeRing)
(Subgroup) edge (Submodule);
%(G1) edge (H1);
%(G2) edge (H2);
\end{tikzpicture}

### 2022-11-07 nakayama/paper.tex
Fixed complaints from nag package
Edits to scientific prose, comments added

%\author{}
\author{ 
	Andrew Cousino \\ {\tt acousino@ku.edu}
	\and 
	Emily E.\ Witt \\ {\tt witt@ku.edu}	
	\and 
	Perry Alexander \\ {\tt palexand@ku.edu}
}	
	%

@@

\emily[inline]{Emily: It might be more consistent to require a ring to contain $1$ in general, instead of adding this to the axioms of a commutative ring.  Perhaps we could do this in the code, and then here, without too much effort? }

@@

\paragraph{Ideals.}
The concept of an {ideal} of a ring can be thought of as an extension of the notion of an integer $x$ in the ring of integers $\mathbb{Z}$. 
An \emph{ideal} of commutative ring $R$ is a subset $I$ of $R$ that is itself 
an abelian group under addition,  
which satisfies the following ``absorption'' property: Given any element $a$ of $I$, 
the product  $x \cdot a$ is again in $I$ for any ring element $x \in R$.

@@

A commutative ring is \emph{local} if it has exactly one maximal ideal.  
Every field is local since the only proper ideal of a field is the zero ideal, though by our observations above, the ring of integers is not local. 
However, the set of all rational numbers that can be written with an odd denominator does form a subring of all rational numbers, and its unique maximal ideal consists of the elements with even numerator; in fact, this ring is the so-called \emph{localization} of $\mathbb{Z}$ at the maximal ideal $2 \mathbb{Z}$.  
The ring of integers modulo $n>1$ is local if and only if $n$ is a power of a prime number $p$,  in which case the unique maximal ideal consists of all multiples of $p$.

@@

\paragraph{A module over a commutative ring.}  Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted $\cdot$, satisfying the following compatibility properties for all $r, s \in R$ and $u, v \in M$.

\begin{enumerate}[leftmargin=6cm,topsep=0cm,itemsep=0cm]
\item  $r \cdot (u + v)  = r \cdot u + r \cdot v$
\item $(r+s) \cdot u = r \cdot u + s \cdot u$
\item $(rs) \cdot u = r \cdot (s \cdot u)$
\item $1 \cdot u = u$ 
\end{enumerate}

From this definition, one can see that a module over a field $F$ is precisely an $F$-vector space, so the notion of a module over an arbitrary commutative ring extends that of a vector space over a field. 
Finitely generated vector spaces form the foundation for matrix algebra, and the extension of this notion to module theory is needed to state Nakayama's Lemma.  We call an $R$-module $M$ \emph{finitely generated} if there exist a fixed finite number of elements $u_1, \ldots, u_n$ of $M$ with the following property:  Given any  $w \in M$, there exist $r_1, \ldots, r_n \in R$ for which  
\[
w = r_1 u_1 + r_1 u_2 + \cdots + r_n u_n.
\]
When $R=F$ is a field and $M=V$ is a finite-dimensional vector space over $F$, one can choose $u_1, \ldots, u_n$ to be a basis for $V$, i.e., $n = \dim V$.  In this case, the choice of scalar coefficients in the expression above for $w \in V$ is unique.  When $R$ is not a field, modules can have more subtle properties, and such expression is typically not unique.

@@

There are alternate statements of Nakayama's Lemma that remove the hypothesis that $R$ must be local.  One can replace the unique maximal ideal with the Jacobson radical of the ring, which is the intersection of all maximal ideals. Alternatively, $I$ is an arbitrary proper ideal of a commutative ring $R$ and  
 $M$ is a finitely generated $R$-module for which $M=I M$, then this ensures the existence of a ring element $r$ congruent to $1$ modulo $I$ such that $rM = 0$, i.e., $r u = 0$ for every $u \in M$.

@@

We start by describing the process of formalization of algebraic structures, and with that in hand, then move on to the formal proof of Nakayama's Lemma.

@@

\begin{figure}[t]
\caption{The hierarchy of our algebraic structures}
\includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
\centering
\todo[inline]{To do:  Check source code to verify whether the hierarchy figure should be altered (subgroups/other subobjects, quotients, finitely generated modules, etc.)}
\end{figure}

We started by defining a semigroup class, which declares a binary operation to
be associative. From here, we build up through monoids, which introduce
identities, to groups, which introduce inverses. Note the double equals ``$==$'' in
these definitions is notation for an arbitrary equivalence relation over the
group's carrier set, which acts as equality.

@@

The next structure defined was a module over a commutative ring, the commutative-algebraic generalization of the notion of a vector
space over a field.

Toward proving 
Here we needed to formalize the notion of linear combinations of
coefficients and module vectors. This was done by dependently typed vectors,
i.e. lists parameterized by their length. Because there is an overload of the
term ``vector'', we will use that term to refer to module vectors, and use the
term ``list'' to mean length-parameterized lists. As we don't use the simpler
kind of lists, this avoids any name collisions. A finitely generated module is
like the vector space \(\mathbf{R}^{n}\) in that there are finitely many
vectors which can generate all other vectors, for \(\mathbf{R}^{n}\) one such
collection of generators are \(\mathbf{e}_{1} = \begin{pmatrix} 1 & 0 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \(\mathbf{e}_{2} = \begin{pmatrix} 0 & 1 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \textellipsis, \(\mathbf{e}_{n} =
  \begin{pmatrix} 0 & 0 & 0 & \cdots & 1\end{pmatrix}^{T}\). In our code,
\texttt{M} is the type of module elements, \texttt{R} is the type of ring
elements which act as coefficients, and \texttt{t A n} is a list whose elements
are of type \texttt{A} and whose length is \texttt{n}.
% \begin{verbatim}
\begin{minted}{coq}
  Definition finitely_generated {n: nat}(basis: t M n) :=
    forall (vector: M),
      exists (coeffs: t R n),
        vector =M= linear_combin coeffs basis.
\end{minted}
% \end{verbatim}

@@

\subsection{Formal Proof of Nakayama's Lemma}

>>>>

%\author{}
\author{ 
	Andrew Cousino \\ {\ttfamily acousino@ku.edu}
	\and 
	Emily E.\ Witt \\ {\ttfamily witt@ku.edu}	
	\and 
	Perry Alexander \\ {\ttfamily palexand@ku.edu}
}	
	%

@@

\emily[inline]{Emily: It might be more consistent to require a ring to contain $1$ in general, instead of adding this to the axioms of a commutative ring. After thinking for a while, I think it is a good idea to make this change in the code, and then here to match.  Drew, does this sound OK?}

@@

\paragraph{Ideals of commutative rings.}
The concept of an {ideal} of a ring can be thought of as an extension of the notion of an integer $x$ in the ring of integers $\mathbb{Z}$. 
An \emph{ideal} of commutative ring $R$ is a subset $I$ of $R$ that is itself 
an abelian group under addition,  
which also satisfies the following ``absorption'' property: Given any element $a$ of $I$, 
the product  $x \cdot a$ is again in $I$ for any ring element $x \in R$.

@@

\paragraph{Local rings.}
A commutative ring is \emph{local} if it has exactly one maximal ideal.  
Every field is local since the only proper ideal of a field is the zero ideal, though by our observations above, the ring of integers is not local. 
However, the set of all rational numbers that can be written with an odd denominator does form a subring of all rational numbers, and its unique maximal ideal consists of the elements with even numerator; in fact, this ring is the so-called \emph{localization} of $\mathbb{Z}$ at the maximal ideal $2 \mathbb{Z}$.  
The ring of integers modulo $n>1$ is local if and only if $n$ is a power of a prime number $p$,  in which case the unique maximal ideal consists of all multiples of $p$.

@@

\paragraph{Module over commutative rings.}  Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted $\cdot$, satisfying the following compatibility properties for all $r, s \in R$ and $u, v \in M$.

\begin{enumerate}[leftmargin=5cm,topsep=0cm,itemsep=0cm]
\item  $r \cdot (u + v)  = r \cdot u + r \cdot v$
\item $(r+s) \cdot u = r \cdot u + s \cdot u$
\item $(rs) \cdot u = r \cdot (s \cdot u)$
\item $1 \cdot u = u$ 
\end{enumerate}

From this definition, one can see that a module over a field $F$ is precisely an $F$-vector space, so the notion of a module over an arbitrary commutative ring extends that of a vector space over a field. 
Finitely generated vector spaces form the foundation for matrix algebra, and the extension of this notion to module theory is needed to state Nakayama's Lemma.  We call an $R$-module $M$ \emph{finitely generated} if there exist a fixed finite number of elements $u_1, \ldots, u_n$ of $M$ with the following property:  Given any  $w \in M$, there exist $r_1, \ldots, r_n \in R$ for which  
\[
w = r_1 u_1 + r_1 u_2 + \cdots + r_n u_n.
\]
The set $\{u_1, \ldots, u_n\}$ is called a \emph{generating set} for the $M$ as an $R$-module.

When $R=F$ is a field and $M=V$ is a finite-dimensional vector space over $F$, one can choose $u_1, \ldots, u_n$ to be a basis for $V$, i.e., $n = \dim V$.  In this case, the choice of scalar coefficients in the expression above for $w \in V$ is unique.  When $R$ is not a field, however, such an expression is typically not unique.

@@

We point out that there are alternate statements of Nakayama's Lemma that do not require the hypothesis that $R$ must be local.  One can replace the unique maximal ideal with the Jacobson radical of the ring, which is the intersection of all maximal ideals. Alternatively, $I$ is an arbitrary proper ideal of a commutative ring $R$ and  
 $M$ is a finitely generated $R$-module for which $M=I M$, then this ensures the existence of a ring element $r$ congruent to $1$ modulo $I$ such that $rM = 0$, i.e., $r u = 0$ for every $u \in M$.

@@

We start by describing our process of formalizing the required algebraic structures detailed in the previous section.  Then, with that in hand, we move on to the formal proof of Nakayama's Lemma.

@@

\begin{figure}[t]
\caption{The hierarchy of our algebraic structures}
\includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
\centering
\todo[inline]{To do:  Check source code to verify whether the hierarchy figure should be altered (subobjects, quotients, finitely generated modules, local rings, etc.)}
\end{figure}

Our foundation begins by defining a semigroup class, which declares a binary operation to
be associative. 
From here, we build up through monoids, which introduce
identities, to groups, which introduce inverses.  
Note the double equals ``$==$'' appearing in
these definitions is notation for an arbitrary equivalence relation over the
group's carrier set, which acts as equality.

@@

\emily[inline]{Emily:  Unfortunately, I think we need to update naming conventions since in mathematics ``vector'' is reserved for an element of a vector space over a field, rather than an element of an arbitrary module, causing confusion.  Similar with ``basis,'' which is especially tricking since two different generating sets of a finitely generated module can have different sizes, but all bases of a vector space have the same cardinality, and an element of a module can typically be written as a scalar combination in multiple ways, while bases are linearly independent.  I've proposed a rewrite of the blue paragraph above it.  Drew, can you check it for accuracy with respect to the code? 
}
\emily[inline]{
We'll also need to update the code.  I think {\tt Vectors} should be something like {\tt ScalarCombinations}. 
}

Finally, we formalized the definition of a module over a commutative ring, the  commutative-algebraic generalization of the notion of a vector
space over a field.  Nakayama's Lemma is a statement about finitely generated modules, and hence we must formalize the notion of a scalar combination of a fixed finite collection of elements $u_1, \ldots, u_n$ of an $R$-module $M$, i.e., expressions of the form  
$r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each $r_i \in R$.

In our formalization of scalar combinations, 
we use  ``list'' to mean length-parameterized lists; since we don't use the simpler
kind of lists, there are no name collisions.
In our code, \texttt{M} is the type of module elements, \texttt{R} is the type of ring
elements,  act as coefficients, and \texttt{t A n} is a list whose elements
are of type \texttt{A} and whose length is \texttt{n}.

\textcolor{blue}{Here we needed to formalize the notion of linear combinations of
coefficients and module vectors. This was done by dependently typed vectors,
i.e. lists parameterized by their length. Because there is an overload of the
term ``vector'', we will use that term to refer to module vectors, and use the
term ``list'' to mean length-parameterized lists. As we don't use the simpler
kind of lists, this avoids any name collisions. A finitely generated module is
like the vector space \(\mathbf{R}^{n}\) in that there are finitely many
vectors which can generate all other vectors, for \(\mathbf{R}^{n}\) one such
collection of generators are \(\mathbf{e}_{1} = \begin{pmatrix} 1 & 0 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \(\mathbf{e}_{2} = \begin{pmatrix} 0 & 1 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \textellipsis, \(\mathbf{e}_{n} =
  \begin{pmatrix} 0 & 0 & 0 & \cdots & 1\end{pmatrix}^{T}\). In our code,
\texttt{M} is the type of module elements, \texttt{R} is the type of ring
elements which act as coefficients, and \texttt{t A n} is a list whose elements
are of type \texttt{A} and whose length is \texttt{n}.
}

\begin{minted}{coq}
  Definition finitely_generated {n: nat}(basis: t M n) :=
    forall (vector: M),
      exists (coeffs: t R n),
        vector =M= linear_combin coeffs basis.
\end{minted}

@@

\subsection{Formally Proving of Nakayama's Lemma}
